Leadership across cultures and contexts: an exploratory analysis

### 2020-05-07 leadership_across_cultures_contexts.Rmd
Worked on discussion (universal section); added feature support plot.

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors; and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable non-predictor terms included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:elastichighstatus)A.

To determine which leader qualities and functions were associated with evidence for high status leaders, we fit an elasticnet logistic model of *High status* as a function of all quality and function variables, (with the lasso penalty, $\alpha = 1$). The non-zero coefficients indicated that *Punishment*, the only identified function, was a negative predictor, whereas
<!-- all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and  -->
*Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several other qualities, were positive predictors. See Figure \@ref(fig:elastichighstatus)B.

@@

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. There were modest or no correlation between features, after aggregating leadership dimensions within features (see Figure \@ref(fig:featurecor)).

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures. See Figure \@ref(fig:varsupport). Leaders benefitted materially, reproductively, and socially in 50% or more of cultures. See Figure \@ref(fig:varsupportcb). Common leader dimensions that did not vary by continental region, subsistence strategy, group context, or leader sex included qualities, such as being knowledgeable and intelligence, wealthly, strategically nepotistic, and generous; and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that there is a fairly high level of evidence for these traits across cultures that does not vary by continental reagion, subsistence strategy, group context, or leader sex. See Table \@ref(tab:leaderuniversals).

Evidence for many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction and prosocial investment; and qualities such as high status and aggressiveness. See Figure \@ref(fig:heatmapdrop1).

Most discussions of leadership in the ethnographic record pertained to male leaders of community-level political groups, with leadership in residential subgroups and kin groups relatively more important in hunter-gatherer and pastoralist societies, and supracommunity political groups more important in horticultural and agricultural societies. Leadership in all social contexts, including economic, military, and religious, was present across all subsistence strategies, however. See Figure \@ref(fig:grouptype). The relative unimportance of supracommunity leadership in hunter-gatherers and pastorlists, and the relative importance of leadership in residential subgroups, highlights a major difference with WEIRD societies that more closely resemble the agricultural societies in our sample that show the opposite pattern.

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. We found evidence that leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). Some common leadership dimensions also did not vary by group context, continental region, subsistence strategy, or leader sex. These included qualities, such as being knowledgeable and intelligence, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that these traits are found in some leaders across all cultures and contexts.

Many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction, and prosocial investment, and qualities such as high status and aggressiveness (see Figure \@ref(fig:heatmapdrop1)).

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent, are experienced, and have a demonstrated history of previous accomplishments or special capacities. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 [p. 236] in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

See Figure \@ref(fig:varsupport). These qualities and functions correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. See the *Synthesis* section below.
<!-- NEED TO DO ADD SECTION OR DELETE HERE -->

@@

>>>>

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_features_support)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

*High status* was the most frequently identified leader quality in the sample, with `r statusPercTxt`% of text records and `r statusPercCult`% of cultures containing evidence that leaders had high social status. To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see supplementary information for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1]`* and *`r rev(highstatus_plot$data$Word)[2]`*, which were the strongest predictive terms; *`r rev(highstatus_plot$data$Word)[3]`* and *`r rev(highstatus_plot$data$Word)[4]`* were also relatively strong predictors; and *`r rev(highstatus_plot$data$Word)[5]`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were also positive predictors. Notable non-predictor terms included, *influence*, *authority*, *headman*, and *chief*. See Figure \@ref(fig:highstatusplot)A.

To determine which leader qualities and functions were associated with evidence for high status leaders, we fit an elasticnet logistic model of *High status* as a function of all quality and function variables, (with the lasso penalty, $\alpha = 1$). The non-zero coefficients indicated that *Punishment*, the only identified function, was a negative predictor, whereas
<!-- all other functions, including *Organize cooperation*, *Ritual functions*, and *Provide counsel/direction* were non-predictors, and  -->
*Wealthy*, *Strategic nepotism*, *Age*, *Experienced/accomplished*, and several other qualities, were positive predictors. See Figure \@ref(fig:highstatusplot)B.

@@

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. Because features were aggregated variables, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

```{r featuressupportplot, fig.width=10, fig.cap="Support of each feature variable at the levels of text records and cultures."}
plot_features_support
```

There were modest or no correlation between features, after aggregating leadership dimensions within features (see Figure \@ref(fig:featurecor)).

@@

## Universal dimensions of leadership

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. We found evidence that leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). Some common leadership dimensions also did not vary by group context, continental region, subsistence strategy, or leader sex. These included qualities, such as being knowledgeable and intelligence, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that there is a fairly high level of evidence for these traits across cultures that does not vary by continental region, subsistence strategy, group context, or leader sex. See Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport).

<!-- Qualities	Knowledgeable/intelligent	81.4 -->
<!-- Functions	Resolve conflict	78.0 -->
<!-- Functions	Misc. social functions	72.9 -->
<!-- Leader Benefits	Material resources	69.5 -->
<!-- Leader Benefits	Increased social status	66.1 -->
<!-- Qualities	Wealthy	66.1 -->
<!-- Functions	Group representative	64.4 -->
<!-- Qualities	Strategic nepotism	64.4 -->
<!-- Qualities	Generosity	61.0 -->

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigous leaders [but see @garfield_investigating_2019]. These dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015], but that model does not emphasize the concrete material benefits seen here. Regarding the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms, the absence of *High status* and *Aggressiveness* on this universal list is notable. Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019].

The following quote, from the Pawnee [@murrayXX], illustrates leadership involving intelligence, conflict resolution, and generosity:

>...a chief was a regulator, not an absolute ruler or tyrant. Although chiefs had considerable authority, their decisions were generally based on a consensus of opinion rather than arbitrary whim. The chief, like his celestial forebear, was supposed to be a guardian of the people, always mindful of their wishes and needs. And even though the office was hereditary in certain families, the man chosen to fill it had to demonstrate humility, generosity, and sagacity, because a jealous or aggressive temperament was considered unbefitting a chief.
<!-- nq18-015 -->

*Coercive authority* is one leadership dimension that is common in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that this dimension was associated with *Punishment*, *Aggressiveness*, and *Feared*, and also with *Control economics*, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies. See Figure \@ref(fig:elasticcoercive). This passage describing Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

>One point in connection with the chief's authority deserves special mention. Power implies not only the possibility of rewarding, but also the means of punishing. This in the Trobriands is as a rule done indirectly, by means of sorcery. The chief has the best sorcerers of the district always at his beck and call. Of course he also has to reward them when they do him a service. If anyone offends him, or trespasses upon his authority, the chief summons the sorcerer, and orders that the culprit shall die by black magic. And here the chief is powerfully helped in achieving his end by the fact that he can do this openly, so that everybody, and the victim himself knows that a sorcerer is after him. As the natives are very deeply and genuinely afraid of sorcery, the feeling of being hunted, of imagining themselves doomed, is in itself enough to doom them in reality. Only in extreme cases, does a chief inflict direct punishment on a culprit. He has one or two hereditary henchmen, whose duty it is to kill the man who has so deeply offended him, that actual death is the only sufficient punishment.

## Variable dimensions of leadership

Evidence for most dimensions of leadership was absent in most cultures: of our `r length(all_study_vars)` dimensions of leadership, there was evidence in less than 50% of cultures for `r sum(df_compare2$Estimate < 0.5)` of them. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb).

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent, are experienced, and have a demonstrated history of previous accomplishments or special capacities. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources.

Evidence for many leadership dimensions did vary across cultures and contexts, however, including functions such as organizing cooperation, providing counsel and direction, and prosocial investment; and qualities such as high status and aggressiveness. See Figure \@ref(fig:heatmapdrop1).

@@

<!-- NEED TO DO ADD SECTION OR DELETE HERE -->

@@

# Notes

## Possible leader quotes (for universal dimensions)

>This headman caters to the needs of the members of smaller group. The system that such heads are always necessarily recognised at wider level, and that, to coordinate, [Page 85] they have a further head over them does not seem to be in existence. The headman, in fact, is primarily meant to look after petty affairs concerning intra-group life; but if need be, he can be used for resolving inter-group disputes. The position of such heads is neither hereditary nor acquired through formal elections. To recognise a head is something spontaneous, especially when some attention-drawing incidents occur. Age factor is not very vital for headmanship. The man who is physically sound and considered active and witty is picked up. The members of the group start considering his views and regard the same as more than those given by any body else. The headman's views carry weight and are attended accordingly.

>At this point, it is appropriate to sum up the role and typical career of a local leader among the Makuna. The political leader ( [unknown]uh[unknown]u ) is, in general, a man who thinks and speaks well. Oratorical skill is one of the basic requirements and [Page 88] qualities of a prestigious leader. Thinking and speaking well mean that he knows how to deal with people, how to prevent social conflict and how to settle disputes and restore social order. The handling of the potentially disruptive, affinal relationships and the sponsoring of communal rituals are the two most important functions of the Makuna leader. Through his oratorical skill-displayed in ceremonial dialogues-disputes are settled and the interests of his followers defended. Through the communal rituals, the social-and ultimately cosmi—corder is maintained or restored.

>The role of the political leader is well defined in Makuna society. It is a specialist role of the same kind as the ritual specialist roles described above. In fact, the role of the political leader is a direct complement to the role of the shaman: while the shaman is a supernatural mediator-a mediator between men and spirits-the political leader is a social mediator-a mediator between men and social groups. The leader represents his group in relation to outsiders of his own kind (human beings), while the shaman deals with non-human “outsiders”, that is, spirits.
>An informant nearly summarized the typical career of a political leader: “A man who wants to become an [unknown]uh[unknown]u must work hard. He must build a large house, cut large gardens and sponsor many dances. He must make a lot of beer and coca and invite many people. He must think good for his people and know how to greet visitors and how to handle enemies. He must be strong and wise.”

>Such headmen, especially the privileged ones, compete vigorously against each other for renown. The privileged headmen or chiefs are loosely inter-graded amongst themselves in ascriptive prestige terms. This fluid scaled model of precedence is not rigid, however, because it is conceptually interlinked with and supplemented by a second model of leadership based on open achievement of renown and name-building through prowess in gardening and ceremonial exchange, wisdom in decision-making and dispute-settlement, and skill in establishing alliances and political support. On these skeletal criteria there is a considerable number of chiefs in Trobriand society at any given point in time. Probably the range 10-25 would quantitatively cover the number of such positio-holders at any one moment over the last eighty years.

@oc06-034 [p. 236] in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders (which also include elements of prestige-based and charismatic leadership discussed above):

>Jape uu me, the brave men, often the tonowis themselves, assumed the leadership during the time of hostilities. They, with other braves and elders, worked out a secret plan of the war strategy, planning the initial attack, deployment of the manpower during the offensive and later on the battle lines, they authorized and often led individual raids into the enemy territory by day or night. If need be, they organized defense of their confederacy’s villages by erecting fortifications of plank fences. In the deployment of their fighting forces they manipulated individual lineages in a similar way as our generals do army divisions. They assigned them their roles in the attack and later their positions on the battle lines.

>On the [Page 118] battle lines they were true leaders, shouting orders and expecting immediate compliance. Warriors disobeying their orders were publicly reprimanded after the battle and put to shame. Those individuals who deserted to the enemy, or otherwise committed treason (usually in-marrying wives) were put to death. Unlike our generals, the jape uu me did not stay protected in the rear, but one usually saw them in the front line and in the lead of an attacking force. By their brave behavior they acted as examples for the rest of the fighting force.
